# **Data Analysis**

***

In this section, we move from data preparation to predictive modeling. Our goal 
is to learn how patterns in DNA methylation across thousands of CpG sites can 
be translated into accurate estimates of chronological age. Because methylation 
data are high-dimensional and biologically structured, no single modeling 
strategy is sufficient on its own. Instead, we hope to build an understanding 
progressively by starting with a strong classical baseline model and then 
introducing more flexible nonlinear approaches. We begin with 
**elastic net regression**, which provides an interpretable and historically 
important foundation for epigenetic clock construction. We then introduce 
**neural networks**, which extend linear models by learning nonlinear 
combinations of CpG features. This leads naturally into **deep learning**, 
where layered architectures allow models to capture increasingly complex 
structure in genomic data. Finally, we discuss 
**overfitting and regularization**, two central concepts that determine whether 
predictive models generalize beyond the datasets on which they were trained. 
Throughout, we hope to provide both the statistical intuition and computational 
tools needed to understand how modern epigenetic clocks are constructed.

## Elastic Net Regression

Before introducing neural networks, we begin with **elastic net regression**, 
as it is central to the history of epigenetic clocks and remains one of the 
most important tools for high-dimensional biological prediction.

Elastic net is a type of *regularized* linear regression model. In traditional 
linear regression, we model an outcome as a weighted sum of predictors:

$$
\text{Age} = \text{Intercept} + \theta_1 \times \text{CpG}_1 + 
\theta_2 \times \text{CpG}_2 + \ldots + \theta_p \times \text{CpG}_p
$$

Each CpG receives a coefficient. Positive coefficients indicate that higher 
methylation at that CpG site is associated with older age, while negative 
coefficients indicate that higher methylation is associated with younger age.

The difficulty is that ordinary regression breaks down when there are more 
predictors than samples, or when predictors are highly correlated. DNA 
methylation data have both problems. Elastic net addresses this by adding a 
*penalty* that discourages the model from becoming too complex.

Elastic net combines two penalties:

- L1 Penalty: Encourages sparsity
- L2 Penalty: Encourages shrinkage and stability

The L1 penalty, also used in lasso regression, can set some coefficients 
exactly to zero. This performs *feature selection* by removing CpG sites that 
do not contribute strongly to prediction.

The L2 penalty, also used in ridge regression, shrinks coefficients toward 
zero without necessarily eliminating them. This is helpful when predictors are 
correlated, because it stabilizes coefficient estimates and prevents the model 
from relying too heavily on any single CpG site.

Elastic net is useful because it balances both behaviors. It can select a 
subset of predictive CpG sites while also handling correlation among
methylation sites.

In the context of epigenetic clocks, elastic net has several advantages. It is 
computationally efficient, relatively interpretable, and well suited to 
settings where the number of CpG sites is much larger than the number of 
samples. As we saw, many foundational epigenetic clocks used elastic net-style 
models for exactly these reasons.

[Add elastic net analysis here]{style="color: red;"}                             <!-- NEED TO ADD! -->

However, elastic net is still a linear model. It assumes that CpG contributions 
are added together in a simple weighted sum. That assumption may be too 
restrictive for biological systems where regulatory mechanisms interact across 
genomic regions, cell states, and developmental pathways.

This limitation motivates our next class of models, neural networks.

## Neural Networks

A **neural network** is a flexible predictive model built from layers of simple 
mathematical transformations. The term *neural network* comes from an analogy 
to the brain: just as biological neurons receive signals from other neurons, 
combine those signals, and pass messages forward through interconnected 
circuits, artificial neural networks pass information through layers of 
computational 'neurons' that transform inputs step by step. Although neural 
networks are often described using this brain metaphor, they are best 
understood as adaptable function *approximators,* that is, mathematical systems 
that learn how to transform input data, such as CpG methylation measurements, 
into useful predictions like chronological age.

In this case study, the input to the neural network is a vector of DNA 
methylation measurements collected across many CpG sites, and the output is a 
single predicted value representing chronological age. Rather than modeling age 
as a simple weighted sum of these measurements, a neural network learns a 
sequence of transformations that gradually reshape the information in the data. 
Each layer combines signals from many CpG sites, adjusts their importance 
through learned weights, and applies a nonlinear transformation that allows 
the model to capture complex biological relationships. By stacking multiple 
layers together, the network can represent patterns that are far more flexible 
than those available to standard linear regression models, making it especially 
well suited for high-dimensional genomic data.

Conceptually:

```{mermaid}
flowchart LR
  A(Methylation Values) --> B(Weighted Combination)
  B --> C(Nonlinear Activation)
  C --> D(Hidden Representation)
  D --> E(Predicted Age)
```

A hidden layer learns an intermediate representation of the data. Instead of 
directly predicting age from CpG sites, the model first transforms the DNA
methylation values into new features. These learned features may capture 
combinations of CpG sites, nonlinear trends, or coordinated methylation 
patterns.

When we stack multiple hidden layers, we create a **deep neural network**. Each 
layer transforms the representation from the previous layer. Early layers may 
learn simpler combinations of CpGs, while later layers may learn more abstract 
patterns useful for age prediction.

[Add illustration here]{style="color: red;"}                                     <!-- NEED TO ADD! -->

For DNA methylation data, this is appealing because aging is unlikely to be 
encoded by independent CpG effects alone. Methylation patterns are shaped by 
chromatin state, cell composition, gene regulation, immune aging, developmental 
programs, and environmental exposures. A neural network can, in principle, 
learn nonlinear relationships among these signals.

However, flexibility comes with tradeoffs. Neural networks can overfit, require 
more computation, and are harder to interpret than elastic net models. They 
also require more modeling decisions: number of layers, number of neurons, 
activation functions, learning rate, dropout, batch size, optimizer, and 
regularization.

## Deep Learning: Why "Deep" Matters

The word **deep** in "deep" learning refers to the use of multiple layers. A 
"shallow" model may transform the input once before prediction. A deep model 
transforms the input repeatedly, allowing each layer to build on the previous 
one.

In image recognition, early layers might learn edges, intermediate layers might 
learn shapes, and later layers might learn object parts. In methylation data, 
the analogy is less visual but still useful. Early layers may learn combinations 
of nearby or correlated CpG sites. Later layers may learn broader patterns 
related to immune composition, developmental regulation, or age-associated 
methylation remodeling.

A deep learning model is trained by minimizing a *loss function*. In this case 
study, the loss is based on absolute prediction error:

$$
\text{Absolute Error} = |\text{True Age} - \text{Predicted Age}|
$$

The model adjusts its weights to reduce this error across the training samples. 
This adjustment is done using an *optimization algorithm* such as Adam.          <!-- Point to a reference here, maybe a margin comment? -->

The training process proceeds iteratively. The model makes predictions, 
calculates errors, updates weights, and repeats. Over many *epochs*, the model 
gradually improves its predictions.

:::{.definition_box}

::::{.definition_box_header}

Important training concepts include:

::::

::::{.inner_block}

**Epoch**: One full pass through the training data.

**Batch Size**: The number of samples used for each weight update.

**Learning Rate**: How large each update step should be.

**Optimizer**: The algorithm used to update model weights.

**Loss Function**: The quantity the model tries to minimize.

**Validation Set**: Held-out data used to monitor performance during training.

**Test Set**: Independent data used for final evaluation.

::::

:::

These concepts are not specific to our motivating methylation data, but central 
to modern deep learning across many different applications. A goal of this case 
study is to make these ideas concrete by applying them to a biologically 
meaningful prediction task.

## Overfitting and Regularization

One of the most important ideas in machine learning is **overfitting**.

A model overfits when it learns patterns that are specific to the training data 
but do not generalize to new data. In genomics, overfitting is especially easy 
because there are many features and often relatively few samples. A model can 
accidentally learn technical artifacts, study-specific batch effects, or random 
noise.

Overfitting can look impressive during training. The model may have very low 
error on the training set. But when applied to new studies, performance may 
collapse.

**Regularization** refers to methods that reduce overfitting by limiting model 
complexity.

Elastic net uses regularization directly through coefficient penalties. Neural 
networks use several forms of regularization, including **dropout**, 
**weight decay**, **early stopping**, and **validation monitoring**.

:::{.definition_box}

::::{.definition_box_header}

Regularization strategies for neural netwroks:

::::

::::{.inner_block}

**Dropout** randomly removes a fraction of hidden units during training. This 
prevents the network from relying too heavily on any one pathway through the 
model. It encourages the network to learn more robust representations.

**Weight decay** penalizes large weights, encouraging smoother and less extreme 
models.

**Cross-validation** repeatedly splits training data into fitting and 
validation subsets, helping evaluate whether model performance is stable.

::::

:::

## Why Compare Elastic Net and Deep Learning?

A central pedagogical theme of this case study is comparison. Deep learning 
should not be introduced as a replacement for statistical modeling. Instead, it 
should be introduced as one part of a modeling toolkit. To understand what deep 
learning contributes, we need meaningful comparisons.

We compare to elastic net for several reasons:

1. First, it is historically central to epigenetic clocks. Many widely used  
clocks were developed using penalized regression.
2. Second, it performs well in high-dimensional settings. A neural network  
that only slightly outperforms a simpler model is not very informative.
3. Third, elastic net is easier to interpret. Its coefficients directly  
indicate which CpG sites contribute to the prediction and in what direction.  
Neural networks are more flexible but less transparent.
4. Fourth, elastic net is computationally efficient. If elastic net performs  
similarly to deep learning, then the added complexity of neural networks may  
not be justified for some applications.

This comparison encourages critical thinking. We do not assume that deep 
learning is universally better. Instead, we ask when additional flexibility 
improves prediction enough to justify increased complexity.

## Introducing Scorcher

Deep learning is often taught using Python frameworks such as TensorFlow,        <!-- NEED REFERENCES/LINKS -->
Keras, or PyTorch. These tools are powerful, but many biomedical researchers, 
statisticians, and data scientists work primarily in R. Switching programming 
languages can create additional barrier and steeper learning curves.

The `scorcher` package is designed to make deep learning more accessible in R 
by building on the `torch` ecosystem while providing a higher-level modeling 
interface. `Scorcher` aims to make neural networks easier to define, inspect, 
train, modify, compare, reuse, and teach. It emphasizes model construction 
through 'tidy' composable functions. This allows users to focus on the 
structure of the model: what the inputs are, what layers are added, what 
activation functions are used, what loss is optimized, and how predictions 
are evaluated.

In this case study, `scorcher` serves two roles:
1. First, it is the computational tool used to build the neural network  
epigenetic clock.
2. Second, it is a teaching framework that makes the deep learning architecture  
visible. Each line in a scorcher pipeline corresponds to a modeling decision.

For example, a model pipeline might say:

```r
model |>
  scorch_layer(linear, in_features = 1000, out_features = 512) |>
  scorch_layer(relu)                                           |>
  scorch_dropout(0.3)                                          |>
  scorch_layer(linear, in_features = 512,  out_features = 512) |>
  scorch_layer(relu)                                           |>
  scorch_dropout(0.3)                                          |>
  scorch_layer(linear, in_featues = 512, out_features = 1)
```

Here, we can see that the model takes 1,000 CpG sites as input, transforms them 
into a 512-dimensional hidden representation, applies a nonlinear (relu) 
activation, uses dropout for regularization, and ultimately returns one 
predicted age.

## A Note on Tidy Function Composition and Piping in R

Many R users are familiar with tidy workflows where data transformations are 
written as a sequence of readable steps. For example:

```r
data |>
  filter(!is.na(age)) |>
  mutate(age_years = as.numeric(age)) |>
  select(sample_id, age_years)
```

This is convenient because it makes analysis pipelines easier to read. Each 
line describes one transformation. The output of one step becomes the input to 
the next. `Scorcher` extends this design philosophy to neural networks. Instead 
of treating model architecture as a hidden object created all at once, we build 
it step by step. This mirrors how we think conceptually:

```text
Start with inputs |>
  Add a dense layer |>
  Add a nonlinear activation |>
  Add regularization |>
  ... Repeat ...
  Add output layer |>
  Compile |>
  Train |>
  Evaluate |>
```

Piping also encourages modularity. If we want to change the model, we can add, 
remove, or modify one step. For example, we can change dropout from 0.3 to 0.5, 
replace the type of activation function, or add another hidden layer. This 
makes experimentation natural, but it also improves reproducibility: a model 
architecture written as a pipeline can be read, reviewed, and discussed. 

[Add scorcher analysis here]{style="color: red;"}                                <!-- NEED TO ADD! -->

***








































If you have been following along but stopped, we could load our wrangled data like so:

```{r}
load(here::here("data", "wrangled", "wrangled_data.rda"))
```

***
<details> <summary> If you skipped the data import section click here. </summary>

An RDA file (stands for R data) of the data can be found [here](https://github.com//opencasestudies/ocs-bp-co2-emissions/tree/master/data/wrangled) or slightly more directly [here](https://raw.githubusercontent.com/opencasestudies/ocs-bp-co2-emissions/master/data/wrangled/wrangled_data.rda). Download this file and then place it in your current working directory within a subdirectory called "wrangled" within a subdirectory called "data" to use the following code. We used an RStudio project and the [`here` package](https://github.com/jennybc/here_here) to navigate to the file more easily.


```{r}
load(here::here("data", "wrangled", "wrangled_data.rda"))
```

</details>
***

:::{.think_question}
::::{.think_question_header}
Question opportunity
::::
::::{.inner_block}
Pose a question about the data analysis.
<details><summary>Answer</summary>
You can use a collapsible details section to provide an answer
</details>
::::
:::

::: {.column-margin}
We might use a column margin note to add an equation or some other annotation for learners or educators.
:::

<!--
Use this section to walk through the methods and code used in the data analysis step-by-step.
Explain how this analysis helps answer the main questions posed earlier in the case study.
-->

***
